Optimal Finite Element Integration

### 2015-08-05 TOMS_Paper/assembly.tex
Several minor fixes

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines for arbitrary multilinear forms. Optimality is defined in terms of floating point operations required to compute a result. We establish a model under which optimality can systematically be reached and comment on the general applicability of this model. The model is integrated in a compiler such that, given a representation of an integration routine, a loop nest in optimal form can automatically be derived. Extensive performance evaluation proves the robustness of the model, showing considerable performance improvements over a number of alternative code generation systems. The effect of low-level optimization is also discussed.
\end{abstract}

@@

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects, which have become increasingly popular over the last years. In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language. The mathematical specification is manipulated by a form compiler, which then generates a representation of assembly operators. By applying these operators to an element in the discretized domain a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed. The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, thus making assembly covering a significant fraction of the overall runtime.

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or "flops", than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific transformation systems or synthesizers. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations that a user should not be expected to write ``by hand''. In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. There is also an on-going effort to produce a novel form compiler, and therefore a novel code generation system for assembly operators, called UFLACS (\cite{Uflacs}).

In spite of a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? With this paper, we aim to make a first step in the direction of solving this problem.

We summarize the contributions of this research as follows
\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
\item We implemented the theory in our compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is the default optimization system in the Firedrake framework. Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
\item We provide an unprecedented (for extension) performance evaluation comparing several code transformation/synthesizers systems in a wide range of forms of increasing complexity, which is essential to demonstrate our optimality claim.
\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we deal with (i.e., optimize) them
\item We enrich the set of low-level transformations described in~\citeN{Luporini}
\end{itemize}

\section{Preliminaries}
\label{sec:background}

%is the computation of contributions of a specific cell in the discretized domain to the linear system which yields the PDE solution. The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector [Olgaard and Wells 2010; AMCG 2010], whose sizes depend on the order of the method. This operation is applied to all cells in the discretized domain. In this work we focus on local matrices, or “element matrices”, which are more costly to compute than element vectors.

We review finite element integration and possible implementations using notation and examples adopted in~\cite{quadrature1} and~\cite{Francis}.

We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are, respectively, a bilinear and a linear form. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are computed in the so called assembly phase. Then, in a subsequent phase, the linear system is solved, usually by means of an iterative method, and $\textbf{u}$ is eventually evaluated.

We focus on the assembly phase, which is often characterized as a two-step procedure: \textit{local} and \textit{global} assembly. Local assembly is the subject of the paper: this is about computing the contributions that an element in the discretized domain provide to the approximated solution of the equation. Global assembly, on the other hand, is the process of suitably ``inserting'' such contributions in $A$ and $b$.

Without loss of generality, we illustrate local assembly in a concrete example; that is, the evaluation of the local element matrix for a Laplacian operator. Consider the weighted Laplace equation
\begin{equation}
- \nabla \cdot (w \nabla u) = 0
\end{equation}
in which $u$ is unknown, while $w$ is prescribed. The bilinear form associated with the weak variational form of the equation is:
\begin{equation}
a(v, u) = \int_\Omega w \nabla v \cdot \nabla u\ \mathrm{d}x
\end{equation}
The domain $\Omega$ of the equation is partitioned into a set of cells (elements) $T$ such that $\bigcup T = \Omega$ and $\bigcap T = \emptyset$. By defining $\lbrace \phi_i^K \rbrace$ as the set of local basis functions spanning $U$ on the element $K$, we can express the local element matrix as
\begin{equation}
\label{stiffness}
A_{ij}^K = \int_K w \nabla \phi_i^K \cdot \nabla \phi_j^K\ \mathrm{d}x
\end{equation}
The local element vector $L$ can be determined in an analogous way. 
%From the computational perspective, its evaluation is however less expensive than that of $A$.

\subsection{Quadrature Mode}
Quadrature schemes are conveniently used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature representation over an element $K$ can be expressed as follows
\begin{equation}
\label{eq:quadrature}
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

@@

\subsection{Tensor Contraction Mode}
Starting from Equation~\ref{eq:quadrature}, exploiting basic mathematical properties we can rewrite the expression as
\begin{equation}
\label{eq:tensor}
A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
\end{equation}
A generalization of this transformation has been proposed in~\cite{Kirby:TC}. By only involving reference element terms, the integral in the equation can be pre-evaluated and stored in a temporary. The evaluation of the local tensor can then be abstracted as
\begin{equation}
A_{ij}^K = \sum_{\alpha} A_{i_1 i_2 \alpha}^0 G_{K}^\alpha
\end{equation}
in which the pre-evaluated "reference tensor" $A_{i_1 i_2 \alpha}$ and the cell-dependent "geometry tensor" $G_{K}^\alpha$ are exposed.

\subsection{Qualitative Comparison}
Depending on the form being considered, the relative performance of the two modes, in terms of number of operations executed, can vary even quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode increasingly more indicate for "complex" forms. On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains "sufficiently small".

These two modes have been implemented in the FEniCS Form Compiler. In this compiler, a simple heuristic is used to choose the most suitable mode for a given form. It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically~\cite{Fenics}. We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}. For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of selecting the optimal mode at the monomials level -- no heuristics, no "global" choice of the mode for all monomials in the form -- without affecting the cost of code generation.

@@

\begin{Def}[Perfect and imperfect loop nests]
A loop nest is said to be \textit{perfect} when non-loop statements appear only in the body of the innermost loop. If this condition does not hold, a loop nest is said \textit{imperfect}. 
\end{Def}

A straightforward property of perfect nests is that hoisting invariant expressions from the innermost loop to the preheader (i.e., the block that precedes the entry point of the nest) is a safe transformation. We will make use of this property.

% In the following, we should represent loops by their iteration variable.

@@

In this work, we are particularly interested in the following class since it naturally arises from the math described in Section~\ref{sec:background}.

@@

Figure~\ref{code:multi_loopnest} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $k$.

@@

\begin{Prop}
\label{prop:share-removal}
Sharing in a perfect multilinear loop nest $LN = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ can always be eliminated.
\end{Prop}
\begin{proof}
The demonstration is by construction and exploits linearity. We want to transform $LN$ into $LN'$ such that there is no sharing is any $L_i \in LN'$. Starting from the innermost loop $L_{i_{n-1}}$, the expressions are ''flattened'' by expanding all products involving terms depending on $L{i_{n-1}}$. Being on the same level of the expression tree, such terms can then be factorized. Due to linearity, each factored product only has one term depending on $L_{i_{n-1}}$, and such term is now unique in the expression. The other terms, independent of $L_{i_{n-1}}$, are, by definition, loop-invariant, and as such can be hoisted at the level of $L_{i_{n-2}}$. This procedure can be applied recursively up to $L_{i_{0}}$: multilinearity allows factorization at each level; perfectness ensures hoisting is always safe.
\end{proof}

This result will be essential to prove that optimality, characterized as follows, can structurally be reached in this class of loops.

@@

A second result follows.
\begin{Prop}
\label{prop:optimal-mln}
A perfect multilinear loop nest $LN$ can always be reduced to optimal form.
\end{Prop}
\begin{proof}
By construction. Loop-dependent terms are logically grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on $L_i$. These sets are sorted in descending order based on their cardinality. By establishing a one-to-one mapping between set indices and loop indices, we produce a new loop permutation. The loop permutation is semantically correct because of perfectness. In this new order, loops are placed such that as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$. Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}. This renders $L_{i_{n-1}}$ optimal. In particular, the number of operations is equal to $LN_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
\end{proof}

@@

The imperfect nest $LN=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear. The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication. We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $LN$ as long as data dependencies are preserved (note that $LN$ is imperfect). One could think of searching for sub-expressions independent of $L_e$, for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count. However, finding or exposing such reducible sub-expressions requires, in general, a full exploration of the expression tree transformation space. This is obviously challenging, and so is the problem of synthesizing an optimal $LN'$ starting from $LN$. Even though we assumed we could generate $LN'$ in non-exponential time, the following issues should be addressed
\begin{itemize}
\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime;
\item the transformations exposing multi-invariant sub-expressions could require expansion and factorization. In terms of number of operations, the save originating from the elimination of the reduction loop could be overwhelmed by the increase in arithmetic complexity of $L_k$ (e.g., expansion can increase the operation count, e.g. $A(B+C) = AB + AC$).
\end{itemize}

We then refine our definition of optimality for generic loop nests as follows

@@

Note how the definition contemplates the possibility for a nest to have multiple innermost loops. In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or result of suitable transformations.

>>>>

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines for arbitrary multilinear forms. Optimality is defined in terms of floating point operations required to compute a result. We establish a model under which optimality can systematically be reached and comment on the general applicability of this model. The model is integrated in a compiler such that, given a representation of an integration routine, a loop nest in optimal form can automatically be derived. Extensive performance evaluation proves the robustness of the model, showing systematic performance improvements over a number of alternative code generation systems. The effect of low-level optimization is also discussed.
\end{abstract}

@@

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects, which have become increasingly popular over the last years. In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language. The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators. By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed. The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, thus making assembly covering a significant fraction of the overall runtime.

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or "flops", than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation systems or synthesizers. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning the on-going effort to produce a novel form compiler in FEniCS, called UFLACS (\cite{Uflacs}), which adds to the abundant set of code generation (or transformation) systems for assembly operators.

However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? With this paper, we aim to make a first step towards solving this problem.

We summarize the contributions of this research as follows
\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
\item We implement the theory in a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is the default optimization system in the Firedrake framework. Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
\item We provide an unprecedented (for its extension) performance evaluation comparing several code generation or transformation systems in a range of forms of increasing complexity. This is essential to demonstrate our optimality claim.
\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we handle them
\item We enrich the set of low-level transformations described in~\citeN{Luporini}
\end{itemize}

\section{Preliminaries}
\label{sec:background}
We review finite element integration and possible implementations using notation and examples adopted in~\cite{quadrature1} and~\cite{Francis}.

We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are, respectively, a bilinear and a linear form. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are "assembled" and subsequently used to solve the linear system through (typically) an iterative method.

We focus on the assembly phase, which is often characterized as a two-step procedure: \textit{local} and \textit{global} assembly. Local assembly is the subject of this article. It consists of computing the contributions of a single element in the discretized domain to the equation solution. In global assembly, such local contributions are "coupled" by suitably inserting them into $A$ and $b$.

We illustrate local assembly in a concrete example, the evaluation of the local element matrix for a Laplacian operator. Consider the weighted Laplace equation
\begin{equation}
- \nabla \cdot (w \nabla u) = 0
\end{equation}
in which $u$ is unknown, while $w$ is prescribed. The bilinear form associated with the weak variational form of the equation is:
\begin{equation}
a(v, u) = \int_\Omega w \nabla v \cdot \nabla u\ \mathrm{d}x
\end{equation}
The domain $\Omega$ of the equation is partitioned into a set of cells (elements) $T$ such that $\bigcup T = \Omega$ and $\bigcap T = \emptyset$. By defining $\lbrace \phi_i^K \rbrace$ as the set of local basis functions spanning $U$ on the element $K$, we can express the local element matrix as
\begin{equation}
\label{stiffness}
A_{ij}^K = \int_K w \nabla \phi_i^K \cdot \nabla \phi_j^K\ \mathrm{d}x
\end{equation}
The local element vector $L$ can be determined in an analogous way. 
%From the computational perspective, its evaluation is however less expensive than that of $A$.

\subsection{Quadrature Mode}
Quadrature schemes are conveniently used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
\begin{equation}
\label{eq:quadrature}
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

@@

\subsection{Tensor Contraction Mode}
Starting from Equation~\ref{eq:quadrature}, exploiting linearity, associativity and distributivity of the involved mathematical operators, we can rewrite the expression as
\begin{equation}
\label{eq:tensor}
A_{ij}^K = \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\alpha_3=1}^n det F_K' w_{\alpha_3} \sum_{\beta=1}^d \frac{X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \int_{K_0} \phi_{\alpha_3} \frac{\partial \phi_{i_1}}{\partial X_{\alpha_1}} \frac{\partial \phi_{i_2}}{\partial X_{\alpha_2}} dX.
\end{equation}
A generalization of this transformation has been proposed in~\cite{Kirby:TC}. By only involving reference element terms, the integral in the equation can be pre-evaluated and stored in temporary variables. The evaluation of the local tensor can then be abstracted as
\begin{equation}
A_{ij}^K = \sum_{\alpha} A_{i_1 i_2 \alpha}^0 G_{K}^\alpha
\end{equation}
in which the pre-evaluated "reference tensor" $A_{i_1 i_2 \alpha}$ and the cell-dependent "geometry tensor" $G_{K}^\alpha$ are exposed.

\subsection{Qualitative Comparison}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for "complex" forms. On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains "sufficiently small". The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-Compiler}). In this compiler, a heuristic is used to choose the most suitable mode for a given form. It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically (\cite{Fenics}). We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}. For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of reasoning at the monomials level -- not at the form level -- without affecting the code generation cost.

@@

\begin{Def}[Perfect and imperfect loop nests]
A loop nest is said to be \textit{perfect} when non-loop statements appear only in the body of the innermost loop. If this condition does not hold, a loop nest is said imperfect. 
\end{Def}

\begin{Def}[Independent basic block]
A sequence of statements that present no data dependencies compose an independent basic block
\end{Def}

We focus on perfect nests whose innermost loop body is an independent basic block. A straightforward property of this class is that hoisting invariant expressions from the innermost to any of the outer loops or the preheader (i.e., the block that precedes the entry point of the nest) is always safe, as long as the obvious loop dependencies are preserved. We will make use of this property. The results of this section could also be generalized to larger classes of loop nests, in which basic block independence does not hold, although this would require a refinement that goes beyond the scope of this paper.

% In the following, we should represent loops by their iteration variable.
We find convenient abstracting mathematical properties at the level of loops

@@

In particular, we are interested in the following class since it naturally arises from the math of finite element integration, reviewed in Section~\ref{sec:background}.

@@

Figure~\ref{code:multi_loopnest} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$.

@@

\begin{Prop}
\label{prop:share-removal}
Assume $LN = [L_{i_{0}}, L_{i_{1}}, ..., L_{i_{n-1}}]$ is a perfect multilinear loop nest with the body of $L_{i_{n-1}}$ being an independent basic block. Then sharing can always be eliminated from $LN$.
\end{Prop}
\begin{proof}
The demonstration is by construction and exploits linearity. We want to transform $LN$ into $LN'$ such that there is no sharing in any $L_i \in LN'$. Starting from the innermost loop $L_{i_{n-1}}$, the expressions are ''flattened'' by expanding all products involving terms depending on $L_{i_{n-1}}$. Being now on the same level of the expression tree, such terms can be factorized. Due to linearity, each factored product only has one term depending on $L_{i_{n-1}}$, and such term is now unique in the expression. The other terms, independent of $L_{i_{n-1}}$, are, by definition, loop-invariant, and as such can be hoisted at the level of $L_{i_{n-2}}$. This procedure can be applied recursively up to $L_{i_{0}}$: multilinearity allows factorization at each level; perfectness ensures hoisting is safe.
\end{proof}

This result will be essential to prove that optimality, characterized as follows, can always be reached in this class of loops.

@@

A second result follows.
\begin{Prop}
\label{prop:optimal-mln}
A perfect multilinear loop nest $LN$ with basic block independence can always be reduced to optimal form.
\end{Prop}
\begin{proof}
By construction. Loop-dependent terms are logically grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on $L_i$. These sets are sorted in descending order based on their cardinality. We produce a new loop permutation that reflects the ${S_i}$ ordering; that is, loops are placed such that, as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$. Note the loop permutation is semantically correct because of perfectness. Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}. This renders $L_{i_{n-1}}$ optimal. In particular, the number of operations is equal to $LN_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
\end{proof}

@@

The imperfect nest $LN=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear. The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication. We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $L_k$ as long as data dependencies are preserved (note that $LN$ is imperfect). One could think of searching for sub-expressions independent of $L_e$, for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count. However, finding or exposing such sub-expressions requires, in general, a full exploration of the expression tree transformation space. This is obviously challenging, and so is the problem of synthesizing an optimal $LN'$ starting from $LN$. Even though we assumed we could generate $LN'$ through a suitable transformation algorithm, the following issues should be addressed
\begin{itemize}
\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing invariant sub-expressions could require expansion and factorization. In terms of number of operations, the save originating from hoisting could be overwhelmed by the increase in arithmetic complexity within $L_k$ (expansion can increase the operation count, e.g. $A(B+C) = AB + AC$).
\end{itemize}

We shall then refine our definition of optimality for generic loop nests as follows

@@

Note how the definition contemplates the possibility for a nest to have multiple innermost loops. In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations.
